---
author: "Graydon Perry"
title: "ps05-rmarkdown-plot"
output: html_document
date: "2023-02-19"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r libraries, include=FALSE}
library(tidyverse)
library(ggplot2)
```

## Load and Check Data

1.) For solving the problems, and answering the questions, create a new rmarkdown document with an appropriate title.

  I created the R-Markdown document.

2.) Load data. How many rows/columns do we have?

```{r load_data, echo=FALSE, results='hide'}
setwd("/Users/graydonperry/Desktop/info201/ps05-rmarkdown-plot")
gapminder <- read_delim("gapminder.csv.bz2")

nrow(gapminder)
ncol(gapminder)
```

  The _gapminder_ dataset has 13055 rows and 25 columns.

3.) Print a small sample of data. Does it look OK?

```{r sample_data, echo=FALSE}
sample_n(gapminder, 5)
```

  Yes, this data looks correct to me.

## Descriptive Statistics

1.) How many countries are there in the dataset? Analyze all three: iso3, iso2 and name.

```{r num_countries, echo=FALSE, results='hide'}
gapminder %>%
  summarize(n = n_distinct(iso3))

gapminder %>%
  summarize(n = n_distinct(iso2))

gapminder %>%
  summarize(n = n_distinct(name))
```

The **iso3** variable shows that there are 253 countries, the **iso2** variable shows that there are 249 countries, and the **name** variables shows that there are 250 countries. 

2a.) Find how many names are there for each iso-2 code. Are there any iso-2 codes that correspond to more than one name? What are these countries?

```{r iso2, echo=FALSE}
gapminder %>%
  group_by(iso2) %>%
  summarize(n = n_distinct(name)) %>%
  arrange(desc(n))

gapminder %>%
  filter(is.na(iso2)) %>%
  summarise(n = unique(name))
```

  Based on my analysis, the **NA** variable is the only code that corresponds
  with multiple names. Those two names are **NA** and **Namibia**.

2b.) Now repeat the same for name and iso3-code. Are there country names that have more than one iso3-code? What are these countries?

```{r iso3, echo=FALSE}
gapminder %>%
  group_by(iso3) %>%
  summarize(n = n_distinct(name)) %>%
  arrange(desc(n))
```

```{r name_iso3, echo=FALSE}
gapminder %>%
  group_by(name) %>%
  summarize(n = n_distinct(iso3)) %>%
  arrange(desc(n))

gapminder %>%
  filter(is.na(name)) %>%
  summarise(n = unique(iso3))
```

  Based on my analysis, the **NA** variable is the only name that corresponds
  with multiple **is03** codes. Those codes are **CHANSIL**, **GBM**, **KOS**,
  and **NLD_CURACO**.

3.)

```{r min_max, echo=FALSE}
gapminder %>%
  filter(!is.na(time)) %>%
  select(time) %>%
  head(1)

gapminder %>%
  filter(!is.na(time)) %>%
  select(time) %>%
  tail(1)

```

  The maximum year is **2019** and the minimum year is **1960**.
  
## CO2 Emissions

1.) How many missing co2 emissions are there for each year? Analyze both missing CO2 and co2_PC. Which years have most missing data?

```{r missing, echo=FALSE}
gapminder %>%
  filter(is.na(co2), is.na(co2_PC), !is.na(time)) %>%
  group_by(time) %>%
  summarize(n = n()) %>%
  arrange(desc(n))
```

  2017, 2018, and 2019 respectively have the most missing CO2 emissions, witch
  217 each.
  
2.) Make a plot of total CO2 emissions over time for the U.S, China, and India. Add a few more countries of your choice. Explain what do you see.

```{r emissions_plot_1, echo=FALSE}
use <-gapminder %>%
  filter(name == "United States of America" | name == "China" | name == "India" |
         name == "Norway" | name == "Italy", !is.na(time), !is.na(co2))

ggplot(data = use, 
       mapping = aes(x = time, 
                     y = co2, 
                     color = name)) +
  geom_point() +
  labs(title = "Total Emissions by Year",
       x = "Year",
       y = "Total Emissions (CO2)")
  
```

  Based on the graph, it appears that **China** and **India** have had
  increasing CO2 emissions since the year 2000, which could reflect increased
  industrialization and overall economic activity in that time. In contrast,
  the **United States** has slightly decreased its emissions since 2000, which
  could reflect changing attitudes on what energy sources should be used, both
  for production and other means. As expected, the two European countries have
  much lower emissions, which most probably indicates their use of alternative
  energy sources and lack of dependence on fossil fuels.
  
3.) Now let’s analyze the CO2 emissions per capita (co2_PC ). Make a similar plot of the same countries. What does this figure suggest?

```{r emissions_plot_2, echo=FALSE}
use <-gapminder %>%
  filter(name == "United States of America" | name == "China" | name == "India" |
         name == "Norway" | name == "Italy", !is.na(time), !is.na(co2_PC))

ggplot(data = use, 
       mapping = aes(x = time, 
                     y = co2_PC, 
                     color = name)) +
  geom_point() +
  labs(title = "Total Emissions by Year",
       x = "Year",
       y = "Total Emissions (CO2 per capita)")
  
```

  In some ways, this graph suggests similar things as the previous, since we
  see some of the same trends. For instance, we still see the general trend  
  that **India** & **China** have increased emissions recently while the  
  **U.S.** has decreased. The main difference in this graph is that, for the
  European countries, we can see how emissions has either grown or decreased as
  a proportion of their population size. Because the trend lines are fairly
  close to the other countries in terms of CO2 per capita, we can infer that
  total CO2 emissions are largely a function of population size.
  
4.) Compute average CO2 emissions per capita across the continents (assume region is the same as continent). Comment what do you see

```{r emissions_continents, echo=FALSE}
gapminder %>%
  filter(!is.na(co2_PC), !is.na(region)) %>%
  group_by(region) %>%
  summarize(n = mean(co2_PC))
```

  Based on the data in the table, it seems like most of the regions have fairly
  similar CO2 emissions  per capita, except for **Africa**, which has a 
  significantly lower average value than the other regions.
  
5.) Make a barplot where you show the previous results–average CO2 emissions per capita across continents in 1960 and 2016.

```{r emissions_barplot, echo=FALSE}
use <- gapminder %>%
  filter(time == 2016 | time == 1960, !is.na(region), !is.na(co2_PC), 
         region != "") %>%
  group_by(region, time) %>%
  summarize(n = mean(co2_PC), .groups = "drop")

ggplot(use, aes(region, n, fill = as.factor(time))) +
  geom_col(position = "dodge") +
  labs(title = "Total Emissions by Region",
       x = "Region",
       y = "Total Emissions (CO2 per capita)") +
  scale_fill_discrete(name = "Year")
```

6.) Which countries are the three largest, and three smallest CO2 emitters (in terms of CO2 per capita) in 2019 for each continent? (Assume region is continent).

```{r 3_large_small, echo=FALSE}
gapminder %>%
  filter(!is.na(co2_PC), time == 2016) %>%
  group_by(region, name) %>%
  summarize(n = mean(co2_PC), .groups = "drop") %>%
  arrange(region, n) %>%
  group_by(region) %>%
  top_n(-3, n)

gapminder %>%
  filter(!is.na(co2_PC), time == 2016) %>%
  group_by(region, name) %>%
  summarize(n = mean(co2_PC), .groups = "drop") %>%
  arrange(region, n) %>%
  group_by(region) %>%
  top_n(3, n)
```

## GDP per capita

1.) Make a scatterplot of GDP per capita versus life expectancy by country, using data for 1960. Make the point size dependent on the country size, and color those according to the continent. Feel free to adjust the plot in other ways to make it better.

```{r gdp_scatterplot_1960, echo=FALSE}
use <- gapminder %>%
  filter(time == 1960, !is.na(lifeExpectancy), !is.na(GDP_PC))

ggplot(data = use, 
       mapping = aes(x = lifeExpectancy, 
                     y = GDP_PC, 
                     color = region)) +
  geom_point(aes(size = totalPopulation)) +
  labs(title = "Life Expectancy vs. GDP per capita",
       x = "Life Expectancy (1960)",
       y = "GDP per capita (1960)") + 
  scale_size_continuous(name = "Total Population") +
  scale_color_discrete(name = "Region")
```

2.) Make a similar plot, but this time use 2019 data only.

```{r gdp_scatterplot_2019, echo=FALSE}
use <- gapminder %>%
  filter(time == 2019, !is.na(lifeExpectancy), !is.na(GDP_PC), !is.na(region))

ggplot(data = use, 
       mapping = aes(x = lifeExpectancy, 
                     y = GDP_PC, 
                     color = region)) +
  geom_point(aes(size = totalPopulation)) +
  labs(title = "Life Expectancy vs. GDP per capita",
       x = "Life Expectancy (2019)",
       y = "GDP per capita (2019)") + 
  scale_size_continuous(name = "Total Population") +
  scale_color_discrete(name = "Region")
```

3.) Compare these two plots and comment what do you see. How has world developed through the last 60 years?

  From both of the plots, we can see that as you raise GDP to a sufficiently 
  high level, life expectancy increases along with it. However, the major 
  difference between the two plots is the scales. For the 1960 plot, life 
  expectancy went up to only 70 while GDP per capita went up to 20000. IN 
  contrast, for the 2019 plot, life expectancy went up to 80 and GDP per capita
  went up to 90000. This tells us that both GDP per capita AND life expectancy
  have increased drastically throughout the world since the beginning of the
  dataset.
  
4.) Compute the average life expectancy for each continent in 1960 and 2019. Do the results fit with what do you see on the figures?

```{r average_le, echo=FALSE}
gapminder %>%
  filter(time == 1960 | time == 2019, !is.na(lifeExpectancy), 
         !is.na(region)) %>%
  group_by(region, time) %>%
  summarize(n = mean(lifeExpectancy), .groups = "drop")
```

  These figures do confirm what we see in the plots above, as average life 
  expectancy has increased from 1960 to 2019 fairly significantly.
  
5.) Compute the average LE growth from 1960-2019 across the continents. Show the results in the order of growth. Explain what do you see.

```{r average_growth, echo=FALSE}
gapminder %>%
  filter(!is.na(lifeExpectancy), !is.na(region), !is.na(time),
         time == 1960 | time == 2019) %>%
  group_by(region, time) %>%
  summarize(avg = mean(lifeExpectancy)) %>%
  mutate(prev = lag(avg, default = 0), growth = avg - prev) %>%
  arrange(desc(growth)) %>%
  filter(time == 2019)
```

  Based on the information, it seems as though Asia and Africa have seen the
  most average growth, while Europe has seen by far the least. Oceania and the
  Americas sit somewhere in the middle.
  
6.) Show the histogram of GDP per capita for years of 1960 and 2019. Try to put both histograms on the same graph, see how well you can do it!

```{r histogram, echo=FALSE}
use <- gapminder %>%
  filter(!is.na(GDP_PC), time == 1960 | time == 2019) 

ggplot(use, aes(GDP_PC, fill = as.factor(time))) +
  geom_histogram(position = "dodge", binwidth = 3500) +
  labs(title = "GDP per capita (by Year)",
       x = "GDP per capita") +
  scale_fill_discrete(name = "Year")
```

7.) What was the ranking of US in terms of life expectancy in 1960 and in 2019? (When counting from top.)

```{r us_rank, echo=FALSE}
gapminder %>%
  filter(!is.na(lifeExpectancy), time == 1960 | time == 2019, name != "") %>%
  group_by(time) %>%
  mutate(rank = rank(-lifeExpectancy), .groups = "drop") %>%
  filter(name == "United States of America") %>%
  select(name, time, rank)
  
```

  The United States of America was ranked 17th in 1960 and 46th in 2019.
  
8.) If you did this correctly, then you noticed that US ranking has been falling quite a bit. But we also have more countries in 2019 – what about the relative rank divided by the corresponding number of countries that have LE data in the corresponding year?

```{r ranking_divided, echo=FALSE}
gapminder %>%
  filter(!is.na(lifeExpectancy), time == 2019, name != "") %>%
  group_by(time) %>%
  mutate(relative_rank = rank(-lifeExpectancy)/n(), .groups = "drop") %>%
  filter(name == "United States of America") %>%
  select(name, time, relative_rank)
```

  The relative rate for the United States of America in 2019, in terms of life 
  expectancy, was 0.235.
  
**Finally**, tell us how many hours did you spend on this PS.

  It took me ~5 hours to comlpete this assignment