---
author: "Graydon Perry"
title: "ps05-rmarkdown-plot"
output: html_document
date: "2023-02-19"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r libraries, include=FALSE}
library(tidyverse)
library(ggplot2)
```

## Load and Check Data

1.) For solving the problems, and answering the questions, create a new rmarkdown document with an appropriate title.

  I Created the R-Markdown Document.

2.) Load data. How many rows/columns do we have?

```{r load_data, echo=FALSE, results='hide'}
setwd("/Users/graydonperry/Desktop/info201/ps05-rmarkdown-plot")
gapminder <- read_delim("gapminder.csv.bz2")

nrow(gapminder)
ncol(gapminder)
```

  The _gapminder_ dataset has 13055 rows and 25 columns

3.) Print a small sample of data. Does it look OK?

```{r sample_data, echo=FALSE}
sample_n(gapminder, 5)
```

  Yes, this data looks correct to me.


## Descriptive Statistics

1.) How many countries are there in the dataset? Analyze all three: iso3, iso2 and name.

```{r num_countries, echo=FALSE, results='hide'}
gapminder %>%
  summarize(n = n_distinct(iso3))

gapminder %>%
  summarize(n = n_distinct(iso2))

gapminder %>%
  summarize(n = n_distinct(name))
```

The **iso3** variable shows that there are 253 countries, the **iso2** variable shows that there are 249 countries, and the **name** variables shows that there are 250 countries. 

2a.)

```{r iso2, echo=FALSE}
gapminder %>%
  group_by(iso2) %>%
  summarize(n = n_distinct(name)) %>%
  arrange(desc(n))

gapminder %>%
  filter(is.na(iso2)) %>%
  summarise(n = unique(name))
```

  Based on my analysis, the **NA** variable is the only code that corresponds with multiple names.
  Those two names are **NA** and **Namibia**.

2b.)

```{r iso3, echo=FALSE}
gapminder %>%
  group_by(iso3) %>%
  summarize(n = n_distinct(name)) %>%
  arrange(desc(n))
```

```{r name_iso3, echo=FALSE}
gapminder %>%
  group_by(name) %>%
  summarize(n = n_distinct(iso3)) %>%
  arrange(desc(n))

gapminder %>%
  filter(is.na(name)) %>%
  summarise(n = unique(iso3))
```

  Based on my analysis, the **NA** variable is the only name that corresponds with multiple **is03**
  codes. Those codes are **CHANSIL**, **GBM**, **KOS**, and **NLD_CURACO**.

3.)

```{r min_max, echo=FALSE}
gapminder %>%
  filter(!is.na(time)) %>%
  select(time) %>%
  head(1)

gapminder %>%
  filter(!is.na(time)) %>%
  select(time) %>%
  tail(1)

```

  The maximum year is 2019 and the minimum yaeer is 1960.
## CO2 Emissions

1.) How many missing co2 emissions are there for each year? Analyze both missing CO2
and co2_PC. Which years have most missing data?

```{r missing, echo=FALSE}
gapminder %>%
  filter(is.na(co2), is.na(co2_PC), !is.na(time)) %>%
  group_by(time) %>%
  summarize(n = n()) %>%
  arrange(desc(n))
```

  2017, 2018, and 2019 respectively have the most missing CO2 emissions, witch 217 each.
  
2.) Make a plot of total CO2 emissions over time for the U.S, China, and India. Add a few
more countries of your choice. Explain what do you see.

```{r emissions_plot_1, echo=FALSE}
use <-gapminder %>%
  filter(name == "United States of America" | name == "China" | name == "India" |
         name == "Norway" | name == "Italy", !is.na(time), !is.na(co2))

ggplot(data = use, 
       mapping = aes(x = time, 
                     y = co2, 
                     color = name)) +
  geom_point() +
  labs(x = "Year",
       y = "Total Emissions (CO2)")
  
```

  Based on the graph, it appears that China and India have had increasing CO2 emissions since the year
  2000, which could reflect increased industrialization and overall economic activity in that time. In
  contrast, the United States has slightly decreased its emissions since 2000, which could reflect
  changing attitudes on what energy sources should be used, both for production and other means. As
  expected, the two European countries have much lower emissions, which most probably indicates their
  use of alternative energy sources and lack of dependence on fossil fuels.
  
3.) Now let’s analyze the CO2 emissions per capita (co2_PC ). Make a similar plot of the
same countries. What does this figure suggest?

```{r emissions_plot_2, echo=FALSE}
use <-gapminder %>%
  filter(name == "United States of America" | name == "China" | name == "India" |
         name == "Norway" | name == "Italy", !is.na(time), !is.na(co2_PC))

ggplot(data = use, 
       mapping = aes(x = time, 
                     y = co2_PC, 
                     color = name)) +
  geom_point() +
  labs(x = "Year",
       y = "Total Emissions (CO2 per capita)")
  
```

  In some ways, this graph suggests similar things as the previous, since we see some of the same
  trends. For instance, we still see the general trend that India & China have increased emissions
  recently while the U.S. has decreased. The main difference in this graph is that, for the European
  countries, we can see how emissions has either grown or decreased as a proportion of their
  population size. Because the trend lines are fairly close to the other countries in terms of CO2 per
  capita, we can infer that total CO2 emissions are largely a function of population size.
  
4.) Compute average CO2 emissions per capita across the continents (assume region is the
same as continent). Comment what do you see

```{r emissions_continents, echo=FALSE}
gapminder %>%
  filter(!is.na(co2_PC)) %>%
  group_by(region) %>%
  summarize(n = mean(co2_PC))
```

  Based on the data in the table, it seems like most of the regions have fairly similar CO2 emissions
  per capita, except for **Africa**, which has a significantly lower average value than the other
  regions.
  
5.) Make a barplot where you show the previous results–average CO2 emissions per capita
across continents in 1960 and 2016.

```{r emissions_barplot, echo=FALSE}
use <- gapminder %>%
  filter(time == 2016 | time == 1960, !is.na(region), !is.na(co2_PC), region != "") %>%
  group_by(region, time) %>%
  summarize(n = mean(co2_PC), .groups = "drop")

ggplot(use, aes(region, n, fill = as.factor(time))) +
  geom_col(position = "dodge") +
  labs(x = "Region",
       y = "Total Emissions (CO2 per capita)") +
  scale_fill_discrete(name = "Year")
```

6.) Which countries are the three largest, and three smallest CO2 emitters (in terms of CO2 per
capita) in 2019 for each continent? (Assume region is continent).

```{r 3_large_small, echo=FALSE}
gapminder %>%
  filter(!is.na(co2_PC), time == 2016) %>%
  group_by(region, name) %>%
  summarize(n = mean(co2_PC), .groups = "drop") %>%
  arrange(region, n) %>%
  group_by(region) %>%
  top_n(-3, n)

gapminder %>%
  filter(!is.na(co2_PC), time == 2016) %>%
  group_by(region, name) %>%
  summarize(n = mean(co2_PC), .groups = "drop") %>%
  arrange(region, n) %>%
  group_by(region) %>%
  top_n(3, n)
```

